### 2018-05-22 write-up/abstract.tex
Finish Implementation

\newpage
{\normalfont\Huge\sffamily\bfseries Abstract}
\vspace{24pt}

This is the abstract. Write a summary of the whole thing. Make 
sure it fits in one page.

>>>>

\newpage
{\normalfont\Huge\sffamily\bfseries TO DO: Abstract}
\vspace{24pt}

In this thesis, I will argue that linear types are an appropriate,
\emph{type-based formalism} for expressing aliasing, read/write permissions,
memory allocation, re-use and deallocation, first, in the context of the APIs
of linear algebra libraries, consequently extending these principles to the
problem of \emph{efficient} matrix-expression compilation.

I will show that framing the problem using linear types can help \emph{reduce
bugs} by making precise and explicit the informal, ad-hoc practices typically
employed by human experts and linear algebra \emph{compilers} and automate
checking them.

As evidence for this argument, I will show programs written with this safety,
precision and explicitness (1) can be just as pleasant and convenient for a
programmer as less efficient, but higher-level linear algebra libraries and (2)
perform just as \emph{efficiently and predictably} as lower-level, less
readable and more error-prone interfaces linear algebra libraries. I will
further show that support for such programs can be implemented \emph{as
libraries} on top of existing programming languages and libraries.

\textbf{Summarise evaluation.}

\textbf{Summary future work.}

### 2018-05-22 write-up/background.tex
Finish Implementation

\subsection{One Too Many Copies and a Thousand Bytes Behind}
\begin{figure}[tbp]
    \centering
    \begin{minted}{python}
# Numpy (Python)
import numpy.matlib 
a = [[1,0],[0,1]] 
b = [[4,1],[2,2]] 
c = numpy.matmul(a,b)
    \end{minted}
    \begin{minted}{julia}
# Julia
c = [1 0; 0 1] * [4 1; 2 2]
    \end{minted}
    \begin{minted}{ocaml}

@@

\begin{figure}[tbp]
    \centering
    \begin{minted}{ocaml}
let mul x y =
  if same_shape x y then
    let y = copy y in
    (_owl_mul (kind x) (numel x) x y y; y)
  else
    broadcast_op (_owl_broadcast_mul (kind x)) x y
    \end{minted}
    \caption{Implementation of Matrix Multiplication in Owl (OCaml).
        Note the `copy' for the result and the unsafe `\_owl\_mul' operation
    used to perform an in-place multiplication.}\label{fig:mat_mul_owl}
\end{figure}

@@

\begin{figure}[tp]
    \begin{minted}{c++}
#include <iostream>
#include <Eigen/Dense>
using namespace std;
int main()
{
    Eigen::Matrix2d a,b,c;
    a << 1, 0, 0, 1; b << 4, 1, 1, 2; c << 0, 0, 0, 0;
    a * b; // new matrix
    c += a * b; // temporary for correctness in case of aliasing
    c.noalias() += a * b; // no temporaries
}
    \end{minted}
    \caption{Some examples of Matrix Multiplication in Eigen. Using expression
        templates (to be discussed later) and \emph{explicitly provided} aliasing
        information, Eigen can emit a single BLAS `dgemm'-like call for the last
	line, mirroring the Fortran example of
	Figure~\ref{fig:fortran_blas}.}\label{fig:cpp_eigen}
\end{figure}

\begin{figure}[tbp]
    \begin{minted}{c++}
noalias(C) = prod(A, B);
// Preferable if T is preallocated
temp_type T = prod(B,C); R = prod(A,T);
prod(A, temp_type(prod(B,C));
prod(A, prod<temp_type>(B,C));
    \end{minted}
    \caption{Boost uBLAS example of Matrix Multiplication. Temporaries need to
        be marked as such to prevent unnecessary re-computation of
        values.}\label{fig:cpp_ublas}
\end{figure}

@@

My proposed solution to this dichotomy is a \emph{domain-specific language}
(DSL), called LT4LA (Linear Types for Linear Algebra). Although for expository
and testing purposes I have defined a concrete-syntax, a full implementation
would make use of a language's syntax-extension features (such as PPX for
OCaml) to \emph{embed} the DSL into the host language. Such an embedding is
straightforward but fairly tedious to implement. As a half-way point, I used
compile-time code generation to make the DSL's output available to OCaml for
testing and evaluation.

@@

If we tried to free one of the read-only aliases instead of
\ltfla{unshare}-ing, then we would get the error show in
Figure~\ref{fig:ltfla_free}. Briefly, a \ltfla{z arr} is a read-write array,
everything else (\ltfla{'x arr} or \ltfla{_ s arr}) is read-only. The types of
\ltfla{free : z arr --o unit}, \ltfla{share : 'x arr --o 'x s arr * 'x s arr}
and \ltfla{unshare : 'x . 'x s arr --o 'x s arr --o 'x arr} are set up so that
you can only free something when you have read-write access to it, guaranteed
by linearity to be the only name in scope with this capability (aliases can
only be read-only).

@@

No exposition of linear types would be complete without a mention of Girard's
Linear Logic~\cite{girard}. As mentioned in the Stanford Encyclopedia of
Philosophy, it is ``a refinement of classical and intuitionistic logic. Instead
of emphasizing truth, as in classical logic, or proof, as in intuitionistic
logic, linear logic emphasizes the role of formulas as resources.'' A walk from
logic to programming along the well-trodden Curry-Howard bridge brings us to
linear types.

@@

I have given an \emph{intuitive} exposition of linear types with
\emph{fractional-capabilities}, emphasising small, but illustrative and
practical code examples, leaving \emph{what} is going on and \emph{why} it
works as details for the next chapter. I have shown that it is possible to
control aliasing and reading/writing capabilities \emph{statically} with a
\emph{type system} -- a programmer can be prevented from writing over and
freeing arrays they do not own whilst being able to safely create read-only
aliases when it makes sense to do so. Unlike Rust's more full-featured
borrow-checker, the next chapter will show that these features can be expressed
and implemented in a concise, \emph{high-level} and language-independent manner
to be packaged up, provided and used \emph{as a library}.

>>>>

\subsection{One Too Many Copies and a Thousand Bytes Behind}
\begin{figure}[tbp]
    \centering
    \begin{minted}[fontsize=\small]{python}
# Numpy (Python)
import numpy.matlib
a = [[1,0],[0,1]]
b = [[4,1],[2,2]]
c = numpy.matmul(a,b)
    \end{minted}
    \begin{minted}[fontsize=\small]{julia}
# Julia
c = [1 0; 0 1] * [4 1; 2 2]
    \end{minted}
    \begin{minted}[fontsize=\small]{ocaml}

@@

\begin{figure}[tbp]
    \centering
    \begin{minted}[fontsize=\small]{ocaml}
let mul x y =
  if same_shape x y then
    let y = copy y in
    (_owl_mul (kind x) (numel x) x y y; y)
  else
    broadcast_op (_owl_broadcast_mul (kind x)) x y
    \end{minted}
    \caption{Implementation of Matrix Multiplication in Owl (OCaml).
        Note the `copy' for the result and the unsafe `\_owl\_mul' operation
    used to perform an in-place multiplication.}\label{fig:mat_mul_owl}
\end{figure}

@@

\begin{figure}[tp]
    \begin{minted}[fontsize=\small]{c++}
#include <iostream>
#include <Eigen/Dense>
using namespace std;
int main()
{
    Eigen::Matrix2d a,b,c;
    a << 1, 0, 0, 1; b << 4, 1, 1, 2; c << 0, 0, 0, 0;
    a * b; // new matrix
    c += a * b; // temporary for correctness in case of aliasing
    c.noalias() += a * b; // no temporaries
}
    \end{minted}
    \caption{Some examples of Matrix Multiplication in Eigen. Using expression
        templates (to be discussed later) and \emph{explicitly provided} aliasing
        information, Eigen can emit a single BLAS `dgemm'-like call for the last
	line, mirroring the Fortran example of
	Figure~\ref{fig:fortran_blas}.}\label{fig:cpp_eigen}
\end{figure}

\begin{figure}[tbp]
    \begin{minted}[fontsize=\small]{c++}
noalias(C) = prod(A, B);
// Preferable if T is preallocated
temp_type T = prod(B,C); R = prod(A,T);
prod(A, temp_type(prod(B,C));
prod(A, prod<temp_type>(B,C));
    \end{minted}
    \caption{Boost uBLAS example of Matrix Multiplication. Temporaries need to
        be marked as such to prevent unnecessary re-computation of
        values.}\label{fig:cpp_ublas}
\end{figure}

@@

My proposed solution to this dichotomy is a \emph{domain-specific language}
(DSL), called LT4LA (Linear Types for Linear Algebra). It offers readable,
explicit management of aliasing, read/write permissions, memory allocation,
re-use and deallocation all with automated checking: offering a safety net to
catch the baby whilst swiftly disposing of the bath water.  Although for
expository and testing purposes I have defined a concrete-syntax, a full
implementation would make use of a language's syntax-extension features (such
as PPX for OCaml) to \emph{embed} the DSL into the host language. Such an
embedding is straightforward but fairly tedious to implement. As a half-way
point, I used compile-time code generation to make the DSL's output available
to OCaml for testing and evaluation.

@@

If we tried to free one of the read-only aliases instead of
\ltfla{unshare}-ing, then we would get the error show in
Figure~\ref{fig:ltfla_free}. Briefly, a \ltfla{z arr} is a read-write array,
everything else (\ltfla{'x arr} or \ltfla{_ s arr}) is read-only. The types of
\ltfla{free}, \ltfla{share} and \ltfla{unshare} (see
Appendix~\ref{chap:primitives}) are set up so that you can only free something
when you have read-write access to it, guaranteed by linearity to be the only
name in scope with this capability (aliases can only be read-only). Conversely,
if we \emph{forgot} to \ltfla{free a}, we would also get \texttt{Variable a not
used} error.

@@

No exposition of linear types would be complete without a mention of Girard's
Linear Logic~\cite{girard}. As mentioned in the Stanford Encyclopedia of
Philosophy, it is ``a refinement of classical and intuitionistic logic. Instead
of emphasizing truth, as in classical logic, or proof, as in intuitionistic
logic, linear logic emphasizes the role of formulas as resources.'' A walk from
logic to programming along the well-trodden Curry-Howard bridge brings us to
linear types~\cite{wadler}.

@@

I have given an \emph{intuitive} exposition of linear types with
\emph{fractional-capabilities}, emphasising small, but illustrative and
practical code examples, leaving \emph{what} is going on and \emph{why} it
works as details for the next chapter. I have shown that it is possible to have
high-level, explicit and automatically checked management of aliasing,
read/write permissions, memory allocation, re-use and deallocation with a
\emph{type system} -- preventing a whole class of errors that can occur with
lower-level languages \emph{at compile time}, using the of context linear
algebra libraries as as a specific example. In the next chapter, I will show
that these features can be expressed and implemented in a language-independent
manner to be provided and used \emph{as a library}.

### 2018-05-22 write-up/implementation.tex
Finish Implementation

\begin{guidance}
This chapter may be called something else\ldots but in general
the idea is that you have one (or a few) ``meat'' chapters which
describe the work you did in technical detail.
\end{guidance}

@@

% Why, runtime representation, trade-offs etc.

To make a linearly-typed language usable, we need some way of using values zero
or more than once, as we would an intuitionistic value. For this, we have in
the type-expressions the !-constructor and in the term-expressions the
\ltfla{Many}-constructor and the \ltfla{let Many <id> = .. in .. } eliminator.
The idea behind the !-type is that the value uses no `resources'
(linearly-typed expressions). To start off with, it is enough to say that
anything which can be passed around by copying, will have a !-type. This
includes integers, elements and booleans. So \ltfla{3 : !int} and \ltfla{3. +.
4. : !elt}. However, all bindings are still linear by default, so to emulate
intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to  \ltfla{let Many x
= <exp> in let Many x = Many (Many x) in <body>} (similarly for function
argument bindings). The reader can check using the rules in the Appendix that
this has the effect of moving \ltfla{x : !t} from the linear to the
intuitionistic environments, only if \ltfla{<exp> : !t}.

@@

Having understood and correctly implemented intuitionism, we now tackle
the problem of how to implement safe aliasing, that is taking a read-write
alias and splitting into read-only aliases, as well as the inverse.

@@

Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
for a positive integer $k$) represents read-only access. So, the \ltfla{'x}
represents a natural number (either a zero \ltfla{z}, variable \ltfla{'x} or a
successor (+1) of a natural number). Hence, you can read from (index) any array
\ltfla{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \ltfla{let !v
<- w[i]} is just sugar for \ltfla{let (w, !v) = get _ w i}). Transparently
rebinding \ltfla{w} with the returned value means a program can appear to use
\ltfla{w} multiple times. The underscore is how a programmer tells the compiler
to automatically \emph{infer} the correct fractional-capability based on the
other arguments passed to the function. In conjunction with the requirement
that functions declarations need type-annotations for their arguments, this
allows a fractional-capability to be correctly inferred in any program.

Fractions exist solely to enable safe aliasing, via the primitive \ltfla{share
: 'x . 'x arr --o 'x s arr * 'x s arr}.  The two arrays returned (which happen
to just be the given array) can now only be read from and not written to.  If
you want to write to this array, you must \ltfla{unshare: 'x .  'x s arr --o
'x s arr --o 'x arr} it with other read-only aliases until you are left with a
value of type \ltfla{z arr}, guaranteeing no other aliases exists.

Given this set-up, we now \emph{statically} have \emph{perfect} information
about aliasing and ownership of values in the program. We can only write to an
array when we own it; ownership guarantees no other aliases exist in scope at
the point of usage. In Figure~\ref{fig:ltfla_kalman}, I show how this perfect
information can be used to write more natural-looking code using value-semantic
expressions which behave in precisely the way we intend it to.  Now the
programmer need not resort to manually tracking and inserting \texttt{noalias}
annotations; instead they can let the loyal and tireless compiler do the heavy
lifting.

@@

\begin{figure}[tp]
    \begin{minted}{ocaml}
let same_resources (wf_a, loc_a) (wf_b, loc_b) =
  let open Let_syntax in
  (* Save state *)
  let%bind {used_vars=prev; env=old_env; _} as state = get in
  (* Reset, run a, save state *)
  let%bind () = put { state with used_vars = empty_used } in
  let%bind res_a = wf_a in
  let%bind {used_vars=used_a; _} as state = get in
  (* Reset, run b, save state *)
  let%bind () = put { state with used_vars = empty_used; env = old_env } in
  let%bind res_b = wf_b in
  let%bind {used_vars=used_b; _} as state = get in
  (* Check if same resources *)
  let keys_a, keys_b = (* convert to (used_a, used_b) to sets *) in
  if Set.equal keys_a keys_b then
    (* merge used_vars and used_b environments *)
  else
    (* report differences *)
    \end{minted}

@@

We have now arrived at the titular \emph{applications} of linear types. I will
show how, in addition to preventing the errors explained hitherto, we can take
linear types one step further and apply it to the domain of efficient
compilation of matrix-expressions.

@@

In contrast, Figure~\ref{fig:ltfla_kalman}, offers the advantages of
\begin{itemize}
    \item aliasing: labelling each step with a different, more meaningful variable name,
    \item easily spotting which resources are being passed in and which are
        allocated for the function (new/copy),
    \item unambiguously seeing \emph{when} and what values are freed;
\end{itemize}
and have the compiler automatically ensure the safety of each of the above by respectively
\begin{itemize}
    \item making it impossible to refer to steps/values which are no longer usable,
    \item ensuring all values are declared and \emph{initialised} correctly before they are used,
    \item checking no values are used after they are freed \emph{or} leaked.
\end{itemize}

@@

\begin{figure}[tb]
    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/kalman.lt}
    \caption{Kalman filter in LT4LA.}\label{fig:ltfla_kalman}
\end{figure}

\subsection{TO DO: Elaboration}

So-called `matrix \emph{expressions}' are still `side-effecting'/consuming
linear variables, and do not produce \emph{values} but a \emph{sequence of
(re-)bindings} which in turn dictate what values are still available/in-scope
after the computation.  As such, compilation cannot be done purely
compositionally via structural-induction on the expression language; it is most
concisely expressed via CPS, where the result of elaborating a matrix
expression is a function that takes as its argument the rest of the computation
expecting to use the result it has just made available under that requested
name.

\section{TO DO: Code Generation}

High-level overview: primitives, translations, optimisations, difficulty, build system.

\subsection{TO DO: Compiling Constructs to OCaml}

\subsection{TO DO: Metaprogramming}

>>>>

% \begin{guidance}
% This chapter may be called something else\ldots but in general
% the idea is that you have one (or a few) ``meat'' chapters which
% describe the work you did in technical detail.
% \end{guidance}

@@

To make a linearly-typed language usable, we need some way of using values zero
or more than once, as we would an intuitionistic value. For this, we have in
the type-expressions the !-constructor and in the term-expressions the
\ltfla{Many}-constructor and the \ltfla{let Many <id> = .. in .. } eliminator.
The idea behind the !-type is that the value uses no `resources'
(linearly-typed expressions). To start off with, it is enough to say that
anything which can be passed around by copying, will have a !-type. This
includes integers, elements and booleans. So \ltfla{3 : !int} and \ltfla{3. +.
4. : !elt}. However, all bindings are still linear by default, so to emulate
intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to  \ltfla{let Many x
= <exp> in let Many x = Many (Many x) in <body>} (similarly for function
argument bindings). The reader can check (using the rules in
Appendix~\ref{chap:ott_spec}) that this has the effect of moving \ltfla{x : !t}
from the linear to the intuitionistic environments, only if \ltfla{<exp> : !t}.

@@

Having started off with linearity and understanding how it helps a programmer
keep track of of memory allocation and deallocation, and then understood and
correctly implemented intuitionism for the only values we wish to be
intuitionistic, and, we now tackle the problem of how to implement safe
aliasing and read/write permissions.

@@

Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
for a positive integer $k$) represents read-only access. So, the \ltfla{'x}
represents a natural number (either a zero \ltfla{z}, variable \ltfla{'x} or a
successor (+1) of a natural number). Hence, you can read from (index) any array
\ltfla{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \ltfla{let !v
<- w[i]} is just sugar for \ltfla{let (w, !v) = get _ w i}). In general, a
left-arrow \ltfla{<-} signifies transparent rebinding with returned values: it
means a program can \emph{appear} to use a variable multiple times, important
for keeping LT4LA usable and readable. The underscore is how a programmer tells
the compiler to automatically \emph{infer} the correct fractional-capability
based on the other arguments passed to the function. In conjunction with the
requirement that functions declarations need type-annotations for their
arguments, this allows a fractional-capability to be correctly inferred in any
program.

Fractions exist to provide both safe aliasing and read/write permissions, via
the primitives \ltfla{share : 'x . 'x arr --o 'x s arr * 'x s arr} and
\ltfla{unshare: 'x .  'x s arr --o 'x s arr --o 'x arr}.  For the former, two
arrays returned (which happen to just be the given array) can now only be read
from and not written to. If you want to write to this array, you must use the
latter to combine other read-only aliases until you are left with a value of
type \ltfla{z arr}, guaranteeing no other aliases exists.

Given this set-up, we now \emph{statically} have \emph{perfect} information
about aliasing and ownership of values in the program. We can only write to or
free an array only when we own it; ownership guarantees no other aliases exist
in scope at the point of usage. In Figure~\ref{fig:ltfla_kalman}, I show how
this perfect information can be used to write more natural-looking code using
value-semantic expressions which behave in precisely the way we intend it to.
Now the programmer need not resort to manually figuring out and inserting
\texttt{noalias} annotations and worrying about what variables can and cannot
be written to or freed; instead they can let the loyal and tireless compiler do
the heavy lifting.

@@

\begin{figure}[tp]
    \begin{minted}[fontsize=\small]{ocaml}
let same_resources (wf_a, loc_a) (wf_b, loc_b) =
  let open Let_syntax in
  (* Save state *)
  let%bind {used_vars=prev; env=old_env; _} as state = get in
  (* Reset, run a, save state *)
  let%bind () = put { state with used_vars = empty_used } in
  let%bind res_a = wf_a in
  let%bind {used_vars=used_a; _} as state = get in
  (* Reset, run b, save state *)
  let%bind () = put { state with used_vars = empty_used; env = old_env } in
  let%bind res_b = wf_b in
  let%bind {used_vars=used_b; _} as state = get in
  (* Check if same resources *)
  let keys_a, keys_b = (* convert to (used_a, used_b) to sets *) in
  if Set.equal keys_a keys_b then
    (* merge used_vars and used_b environments *)
  else
    (* report differences *)
    \end{minted}

@@

We have now arrived at an extended application of linear types. I will show
how, we can apply the ability to automatically check aliasing, read/write
permissions, memory allocation, re-use and deallocation, to the domain of
matrix-expression compilation.

@@

In contrast, Figure~\ref{fig:ltfla_kalman}, offers the advantages of
\begin{itemize}
    \item aliasing: labelling each step with a different, more meaningful variable name,
    \item easily spotting which resources are being passed in and which are
        allocated for the function (new/copy),
    \item unambiguously seeing \emph{when} and what values are freed;
\end{itemize}
and have the compiler automatically ensure the safety of each of the above by respectively
\begin{itemize}
    \item making it impossible to refer to steps/values which are no longer usable,
    \item ensuring all values are declared and \emph{initialised} correctly before they are used,
    \item checking values are neither used after they are freed \emph{nor} leaked.
\end{itemize}

@@

\begin{figure}[tp]
    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/kalman.lt}
    \caption{Kalman filter in LT4LA.}\label{fig:ltfla_kalman}
\end{figure}

\subsection{Elaboration}

All of the syntax in Figure~\ref{fig:ltfla_kalman} can be unambiguously
desugared \emph{before} type-checking, through fairly simple pattern matching.
Matrix expression compilation is well-trodden territory in
academia~\cite{rocklin_thesis, fabregat_thesis, gunnels_flame, linnea, taco}
but this is, to my knowledge, the first type-based approach to it.

An overview of the translations are in Figure~\ref{fig:mat_patterns}; details
about choosing between `symm' or `gemm' are omitted for brevity. Before
settling on this approach, I tried implementing a more general type-directed,
nested matrix expression compiler; I will now highlight some of the
difficulties inherent in the problem.

\begin{figure}[tp]
    \begin{minted}[fontsize=\small]{ocaml}
let x <- [| y |] in ..              ==> let (y,x) = copyM_to _ y x in ..
let x <- new [| y |] in ..          ==> let (y,x) = copyM _ y in ..
let x <- [| a*b + c |] in ..        ==> let x <- [| 1.*a*b + 1.*c |] in ..
let x <- [| i.*a*b + j.*c |] in ..  ==> let ((a,b), c) = (* BLAS *) in ..
let x <- new (m, n) [| a*b |] in .. ==> let c = matrix m n in
                                        let x <- [| 1.*a*b + 0.*c |] in ..
    \end{minted}
    \caption{Syntactic translations of matrix expressions to linearly-typed
        matrix functions. Further annotations on the matrix variables
        (\ltfla{sym} or \ltfla{^T}) determine which BLAS routine and what
        parameters to call it with.}\label{fig:mat_patterns}

One of the first hurdles I encountered was compositionality: to compile $AB +
C$ it is typically better to use a BLAS routine directly rather than but first
compiling $A$, then $B$, then adding a call to multiply them, then $C$ and
finishing with a call to add the results.

Another compositionality problem is that a call to a linear function does not
just return a result, but a sequence of re-bindings that dictate which
variables are still in scope/usable.  As such, to compile an expression, you
need to provide a CPS-style function of type $var \rightarrow exp$ representing
how you would use a variable representing the result in the rest of the
expression.

However, the type of that variable also determines how you can use it: can you
write to it or must you copy it? This information depends on how the expression
representing the variable was elaborated and adds to the complexity of the
pattern-matching.

Copying leads us into dealing with temporaries: do you first allocate all
temporaries in new matrices and then analyse lifetimes and dimensions to figure
out which slots of memory can be re-used? Or do you try and infer the available
resources from the environment as you go?  Can, and should, you type arrays and
matrices (or $n$-dimensional tensors) as the same?

Adding in more and more considerations, the problem starts to resemble register
allocation: there are registers of different types and sizes, many instructions
to choose from, a cost model to take into account all whilst trying to balance
the number of registers in use and the number of instructions emitted.

\section{Code Generation}

Code generation is a thankfully straightforward mapping from core LT4LA
constructs to OCaml constructs, with the appropriate sprinkling of \ltfla{Many}
constructors for integer, element and boolean literals.

To make the code produced readable, I added a few `optimisations', to the
compiler, to `re-sugar' some of the constructs translated where appropriate.
Almost all of them involved the erasing the !-eliminator which is not needed in
regular, intuitionistic OCaml. So, so we can simply replace any expression of
the form:

\begin{itemize}

    \item \ltfla{let xy, z = <exp> in let x,y = xy in <body>} with
        \ltfla{let (x,y), z = <exp in <body>},

    \item \ltfla{let Many x = x in let Many x = Many (Many x) in <body>} with
        \ltfla{<body>},

    \item \ltfla{let Many x = <exp> in let Many x = Many (Many x) in <body>} with
        \ltfla{let x = <exp> in <body>},

    \item \ltfla{let Many x = Many <exp> in <body>} with \ltfla{let x = <exp> in <body},

    \item \ltfla{let Many f = fix (f, x, .., <exp>, ..) in <body>} with
        \ltfla{let rec f x = <exp> in <body>}.

\end{itemize}

The end result is visible in Figures~\ref{fig:ocaml_sumarray}
and~\ref{fig:ocaml_kalman}: OCaml code that is not obviously safe and correct
with respect to linearity, as promised at the start of the chapter.

\begin{figure}[tp]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let rec f i n x0 row =
  if Prim.extract @@ Prim.eqI i n then (row, x0)
  else
    let row, x1 = Prim.get row i in
    f (Prim.addI i (Many 1)) n (Prim.addE x0 x1) row
in
f
    \end{minted}
    \caption{Recursive OCaml function for a summing an array, generated from
        Figure~\ref{fig:ltfla_sumarray}, passed through \texttt{ocamlformat} for
        presentation.}\label{fig:ocaml_sumarray}

Even though Figure~\ref{fig:ocaml_kalman} is in OCaml, I see little reason that
it cannot be C or Fortran. Translating LT4LA's core language constructs to
those languages would not be as easy as translating to OCaml. However, it is
easy to see that both Figures~\ref{fig:fortran_kalman}
and~\ref{fig:ocaml_kalman} are realisations of a concise, linearly-typed Kalman
filter~\ref{fig:ltfla_kalman} \emph{specification} that describes the whole
behaviour and intent (with respect to read/writes permissions, memory
management and aliasing) of the program and the BLAS primitives it uses more
accurately than a Fortran, C or OCaml implementation could.

\begin{figure}[p]
    \begin{minted}[linenos, fontsize=\footnotesize]{ocaml}
let kalman sigma h mu r_1 data_1 =
  let h, _p_k_n_p_ = Prim.size_mat h in
  let k, n = _p_k_n_p_ in
  let sigma_h = Prim.matrix k n in
  let (sigma, h), sigma_h =
    Prim.symm (Many true) (Many 1.) sigma h (Many 0.) sigma_h
  in
  let (sigma_h, h), r_2 =
    Prim.gemm (Many 1.) (sigma_h, Many false) (h, Many true) (Many 1.) r_1
  in
  let (h, mu), data_2 =
    Prim.gemm (Many 1.) (h, Many false) (mu, Many false) (Many (-1.)) data_1
  in
  let h, new_h = Prim.copy_mat_to h sigma_h in
  let r_2, new_r = Prim.copy_mat r_2 in
  let chol_r, sol_h = Prim.posv new_r new_h in
  let chol_r, sol_data = Prim.potrs chol_r data_2 in
  let () = Prim.free_mat chol_r in
  let h_sol_h = Prim.matrix n n in
  let (h, sol_h), h_sol_h =
    Prim.gemm (Many 1.) (h, Many true) (sol_h, Many false) (Many 0.) h_sol_h
  in
  let () = Prim.free_mat sol_h in
  let h_sol_data = Prim.matrix n (Many 1) in
  let (h, sol_data), h_sol_data =
    Prim.gemm (Many 1.) (h, Many true) (sol_data, Many false) (Many 0.) h_sol_data
  in
  let mu, mu_copy = Prim.copy_mat mu in
  let (sigma, h_sol_data), new_mu =
    Prim.symm (Many false) (Many 1.) sigma h_sol_data (Many 1.) mu_copy
  in
  let () = Prim.free_mat h_sol_data in
  let h_sol_h_sigma = Prim.matrix n n in
  let (sigma, h_sol_h), h_sol_h_sigma =
    Prim.symm (Many true) (Many 1.) sigma h_sol_h (Many 0.) h_sol_h_sigma
  in
  let sigma, sigma_copy = Prim.copy_mat_to sigma h_sol_h in
  let (sigma, h_sol_h_sigma), new_sigma =
    Prim.symm (Many false) (Many (-1.)) sigma h_sol_h_sigma (Many 1.) sigma_copy
  in
  let () = Prim.free_mat h_sol_h_sigma in
  ((sigma, (h, (mu, (r_2, sol_data)))), (new_mu, new_sigma)) )
in
kalman
    \end{minted}
    \caption{OCaml code for a Kalman filter, generated from
        Figure~\ref{fig:ltfla_kalman}, passed through \texttt{ocamlformat} for
        presentation.}\label{fig:ocaml_kalman}

\subsection{Build System}

So once you have written your memory-optimised program with all the features
and support provided by LT4LA and made it produce well-typed, compilable OCaml
code, the question then becomes, how to use this code. This process will vary
across ecosystems, but within OCaml, the new build-system on the block
\emph{Dune} has support for generating, compiling and linking OCaml modules at
\emph{compile time}. Indeed, this is how I have written tests and benchmarks
for programs produced by LT4LA from within OCaml. In particular, I can use my
linearly-typed Kalman filter implementation just like and with any other OCaml
function.

Advantages of generating code at compile time include avoiding runtime overhead
as well as catching any interface changes between the generated code and the
code that uses it via OCaml's type system (another benefit of embedding LT4LA's
type system inside of OCaml).

I suspect that this approach will be very valuable to not just users of
libraries such as Numpy or Owl, but also their implementors: library functions
which presented a safe, value-semantic interface but use unsafe, mutating
operations on the inside could now be expressed using LT4LA and have an extra
safety-net and automatic checking for their implementations.

\section{Summary}

I have explained how a few core features -- linearity, the \ltfla{Many}
constructor, value-restriction, fractional-capabilities with inference,
if-expressions and recursive functions -- are enough to \emph{statically
capture and automatically check} aliasing, read/write permissions, memory
allocation, re-use and deallocation of real world linear algebra library
routines. I have also demonstrated that simple pattern-matching and desugaring
provides the potential for a new, \emph{type-directed} approach to matrix
expression compilation. Lastly I have shown that it is possible to use these
features with \emph{existing} languages and frameworks.\footnote{As mentioned
in the previous chapter, if the host language supports
\emph{syntax-extensions}, like PPX for OCaml, it would be possible to use LT4LA
features \emph{from within} OCaml instead of just alongside it.}

### 2018-05-22 write-up/introduction.tex
Finish Implementation

\begin{guidance}
    This is the introduction where you should introduce your work.  In
    general the thing to aim for here is to describe a little bit of the
    context for your work --- why did you do it (motivation), what was the
    hoped-for outcome (aims) --- as well as trying to give a brief
    overview of what you actually did.

    It's often useful to bring forward some ``highlights'' into
    this chapter (e.g.\ some particularly compelling results, or
    a particularly interesting finding).

    It's also traditional to give an outline of the rest of the
    document, although without care this can appear formulaic
    and tedious. Your call.
\end{guidance}

>>>>

% \begin{guidance}
%     This is the introduction where you should introduce your work.  In
%     general the thing to aim for here is to describe a little bit of the
%     context for your work --- why did you do it (motivation), what was the
%     hoped-for outcome (aims) --- as well as trying to give a brief
%     overview of what you actually did.
% 
%     It's often useful to bring forward some ``highlights'' into
%     this chapter (e.g.\ some particularly compelling results, or
%     a particularly interesting finding).
% 
%     It's also traditional to give an outline of the rest of the
%     document, although without care this can appear formulaic
%     and tedious. Your call.
% \end{guidance}

\prechapter{%
}%

### 2018-05-22 write-up/related_work.tex
Finish Implementation

\prechapter{%
    Now that I have described my contributions, I will explain how it relates
    to existing work, leaving brief discussions on future work to the next
    chapter. I strongly believe matrix expression compilation research would
    benefit greatly from a comprehensive literature review but unfortunately
    that is beyond the scope of this chapter.
}%

@@

MetaOCaml~\cite{metaocaml} and Scala with Lightweight Modular
Staging~\cite{scala_lms} are systems which support multi-stage programming. A
typical example of this is the generation of Fast Fourier-Transform kernels,
specialised to a desired array length.

>>>>

\prechapter{%
    Now that I have described my contributions, I will explain how it relates
    to existing work, leaving brief discussions on future work to the next
    chapter. I strongly believe the field of matrix expression compilation would
    benefit greatly from a comprehensive literature review but unfortunately
    that is beyond the scope of this chapter.
}%

@@

MetaOCaml~\cite{metaocaml} and Scala with Lightweight Modular
Staging~\cite{scala_lms} are systems which support multi-stage programming. A
typical example of this is the generation of Fast Fourier-Transform kernels,
specialised to a desired array length. Combining this with recent work on
generalising and automatically deriving \emph{partially static} representations
of data~\cite{yallop}, it may be possible to apply such techniques to
\emph{tensor algebra} expressions.
